The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2019-11-02 categorical-zero-one.tex
Minor fixes and some more comments

\emph{Markov categories} are an approach to probability and statistics based on category theory, developed by the first author in \cite{markov_cats}, building on much previous work by other authors (see \Cref{background}).
The basic idea is to axiomatize the structure and properties of ``categories of stochastic processes'' - for instance, the category $\Stoch$ where objects are measurable spaces and morphisms are Markov kernels (hence the name).
The goal is to state and prove results from probability and statistics in this context.
This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}).
Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry - defining certain basic relations between the geometric objects at hand, and setting down axioms about them - with the more modern description of geometric objects as (very roughly) certain subsets of $\bR^n$.

@@

Very roughly, the Kolmogorov zero--one law states that, given an infinite collection of independent random variables, and an event which is in the $\sigma$-algebra generated by the variables, which is independent of any finite subset of variables, the event has probability one or zero.

@@

A symmetric monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently, it is a monoidal category equipped with morphisms
\[
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\]
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. Here are the main examples that we will be considering.

@@

In $\FinStoch$, the deterministic morphisms $X \to Y$ are exactly the $\{0,1\}$-valued stochastic matrices. Since these can be identified with honest functions $X \to Y$, we conclude that $\FinStoch_\detc$ is equivalent to the category of finite sets. In $\Stoch$, the deterministic morphisms $f : X \to Y$ are exactly those Markov kernels for which $f(S|x) \in \{0,1\}$ for every $S \in \Sigma_Y$ and $x \in X$. In other words, as the term suggests, the deterministic morphisms are those which do not involve any randomness. In $\CRing_+\op$, the deterministic morphisms $R \to S$ are precisely those morphisms which are represented by additive unital maps $S \to R$, i.e.~the ring homomorphisms. Thus the subcategory of deterministic morphisms is exactly the opposite of the usual category of commutative rings.

@@

\begin{definition}
	Given a morphism $p : A \to X_1 \otimes \ldots \otimes X_n$ in a Markov category, we say that $p$ \emph{displays the conditional independence} $X_1 \perp \ldots X_n \mid\mid A$ if the equation
	\[
		\tikzfig{cond_ind}
	\]
	holds.
\end{definition}

@@

In particular, this proves that \Cref{defn_kolmogorov_ext} determines the comonoid structure on a Kolmogorov product uniquely: $\cop{\bigotimes_j X_j}$ has to be the diagonal morphism in $\cC_\detc$, and this condition determines $\bigotimes_j X_j$ up to \emph{deterministic} isomorphism.

\begin{example}
    \label{setmulti}
    Let $\SetMulti$ be the category with objects ordinary sets, maps $f: X \to Y$ given by functions (i.e maps in the ordinary category of sets) $f: X \to \cP(Y)\setminus \emptyset$, and composition given by
    \[f\circ g(x) = \bigcup_{y \in g(x)} f(y)\]
    for $g: X \to Y, f: Y \to Z$.
    We think of the morphisms in $\SetMulti$ as \emph{multivalued functions}.
    See also \cite[Example~2.6]{markov_cats}.
    The Cartesian product of sets gives it a symmetric monoidal structure, if we set \[(f \tensor g)(x_1,x_2) = f(x_1) \times g(x_2) \subset Y_1 \times Y_2,\] for $f:X_1 \to Y_1, g: X_2 \to Y_2$.
    This is a semicartesian monoidal structure, and we can give it a Markov category structure by setting
    $\cop{X}(x) = \{(x,x)\} \subset X \times X$.
    Then the deterministic morphisms are precisely those sending each point into a singleton set, and there is an obvious isomorphism of categories
    $\SetMulti_\detc \cong \sf{Set}$.
    Of course, $\sf{Set}$ actually has all infinite products, but they do not form Kolmogorov products, or even infinite tensor products in $\sf{SetMulti}$.
    To see this, consider the family $X_n = \{0,1\}$ for $n \in \bN$.
    A map $* \to \prod_{n\in \bN} \{0,1\} = X_\bN$ is just a subset of this infinite product.
    Given such a subset $U$, the finite marginals $* \to X_F$ for $F \subset \bN$ are the images of $U$ under the projection map.
    But these images do not determine $U$. For instance, consider the subset $U$ of those sequences containing at least one $1$.
    The image in $X_F$ is all of $X_F$ (since any finite sequence of $0$s and $1$s can be extended to one containing a $1$).
    But $U$ is not all of $X_\bN$.

@@

\begin{lemma}
	\label{aseqlemma}
	Let $\cC$ be a causal Markov category. Let $p : A \to X$ and $f,g: X \to Y$ be morphisms in $\cC$ with $f$ deterministic and such that
	\[
		\tikzfig{aseqlemma}
	\]
	Then $f =_{p\as} g$ (\Cref{defnasequal}).
\end{lemma}
\begin{proof}
    Consider the morphism
    \[
	    \tikzfig{aseqlemma1}
    \]
    Applying the assumed equation together with determinism of $f$ shows that marginalizing over the first output gives the same morphism $A \to Y \otimes Y$ as marginalizing over the second output. Therefore, applying the causality property of \Cref{causal_defn} with $h_1$ and $h_2$ the two marginalizations, we can conclude the equality
    \[
	\tikzfig{aseqlemma2}
    \]
    Marginalizing this further over the middle output and truncating the redundant pieces proves the desired equation.
\end{proof}

@@

\todo[inline]{E: Not sure these examples are that great.. it seems like we do an inordinate amount of work to get the infinite tensor products, compared to the results. What do you think?}

@@

\begin{definition}
    Consider the hyperspace monad on $\Top$ defined in \cite[Definition~2.4]{hyperspace}.
    The Kleisli category associated to this monad is denoted $\Kl(H)$.
    Using \cite[Corollary~2.53]{hyperspace} and \cite[Proposition~3.1]{markov_cats}, this is a Markov category.
    Concretely, this category has the following structure:
    \begin{enumerate}
        \item The objects are topological spaces.
        \item A map $f: X \to Y$ consists of a continuous function $X \to HY$, where $HY$ denotes the set of closed subsets, equipped with the topology generated by subbasic opens of the form $\{V \subset Y \text{ closed} \mid V \cap U \neq \emptyset\}$ for $U \subset Y$ open.
        \item Composition is defined as in $\SetMulti$, by the closure of the union.
        \item The monoidal product is the Cartesian product of topological spaces, and $(f \tensor g)(x_1,x_2) = f(x_1) \times g(x_2)$.
        \item The comultiplication is defined by $\cop{X}(x) = \cl(\{(x,x)\})$
    \end{enumerate}
\end{definition}
\begin{proposition}
    $\Kl(H)$ has Kolmogorov products of any cardinality, given simply by the ordinary infinite product of topological spaces in the product topology.
\end{proposition}
\begin{proof}
    First we observe that $H(\prod_{j\in J} X_j)$ is the cofiltered limit of the spaces
    $H(\prod_{j\in F} X_j)$.
    To see this, note that the marginalizations induce a continuous map
    \[\alpha: H(\prod_{j\in J} X_j) \to \lim_{F \subset J} H(\prod_{j \in J} X_j)\]
    This is an injection - given distinct closed subsets $V, W \subset \prod X_j$, and let $x \in V \setminus W$.
    Then let $U \ni x$ be an open neighborhood of $x$ not intersecting $W$.
    Since $U$ is open in the product topology, we can find some finite subset $F \subset J$ so that 
    \[y = (y_j) \in U \Leftrightarrow y_j \in \pi_j(U) \text{ for $j\in F$}\]
    Then $\pi_F(x) \in \pi_F(U)$, which is disjoint from $\pi_F(W)$. Hence $\alpha(V) \neq \alpha(W)$.

>>>>

\emph{Markov categories} are an approach to probability and statistics based on category theory, developed by the first author in \cite{markov_cats}, building on much previous work by other authors (see \Cref{background}).
The basic idea is to axiomatize the structure and properties of ``categories of stochastic processes'' - for instance, the category $\Stoch$ where objects are measurable spaces and morphisms are Markov kernels (hence the name).
The goal is to state and prove results from probability and statistics in this context.
This represents a \emph{synthetic} approach to probability, in contrast with the usual measure-theoretic approach (which can be called \emph{analytic}).
Instead of defining \emph{what} the words probability, distribution, random variable, etc, \emph{mean}, we instead state by which operations these objects can be combined and related to each other, and which properties they are supposed to satisfy.
We can draw an analogy to the Euclidean approach to geometry - defining certain basic relations between the geometric objects at hand, and setting down axioms about them - with the more modern description of geometric objects as (very roughly) certain subsets of $\bR^n$.\improve{The explanation of this analogy is a bit wonky}

@@

Very roughly, the Kolmogorov zero--one law states that, given an infinite collection of independent random variables, and an event which is in the $\sigma$-algebra generated by the variables, which is independent of any finite subset of variables, the event has probability one or zero.\improve{add references for the Kolmogorov and Hewitt--Savage 01 laws}

@@

A symmetric monoidal category $\cC$ is \emph{semicartesian} if the monoidal unit $I \in \cC$ is terminal. Equivalently\todo{Should there be a reference to this equivalence?}, it is a monoidal category equipped with morphisms
\[
	X \otimes Y \longrightarrow X, \qquad X \otimes Y \longrightarrow Y,
\]
which are natural in $X$ and $Y$ and coincide with the monoidal structure isomorphisms whenever $X = I$ or $Y = I$. Here are the main examples that we will be considering.

@@

In $\FinStoch$, the deterministic morphisms $X \to Y$ are exactly the $\{0,1\}$-valued stochastic matrices. Since these can be identified with honest functions $X \to Y$, we conclude that $\FinStoch_\detc$ is equivalent to the category of finite sets. In $\Stoch$, the deterministic morphisms $f : X \to Y$ are exactly those Markov kernels for which $f(S|x) \in \{0,1\}$ for every $S \in \Sigma_Y$ and $x \in X$. In other words, as the term suggests, the deterministic morphisms are those which do not involve any randomness. In $\CRing_+\op$, the deterministic morphisms $R \to S$ are precisely those morphisms which are represented by additive\todo{should this be ``multiplicative''?}  unital maps $S \to R$, i.e.~the ring homomorphisms. Thus the subcategory of deterministic morphisms is exactly the opposite of the usual category of commutative rings.

@@

\begin{definition}
	Given a morphism $p : A \to X_1 \otimes \ldots \otimes X_n$ in a Markov category, we say that $p$ \emph{displays the conditional independence} $X_1 \perp \ldots \perp X_n \mid\mid A$ if the equation
	\[
		\tikzfig{cond_ind}
	\]
	holds.
\end{definition}

@@

In particular, this proves that \Cref{defn_kolmogorov_ext} determines the comonoid structure on a Kolmogorov product uniquely: $\cop{\bigotimes_j X_j}$ has to be the diagonal morphism in $\cC_\detc$, and \Cref{kolmprod_catdet} determines $\bigotimes_j X_j$ up to \emph{deterministic} isomorphism.

\begin{example}
    \label{setmulti}
    Let $\SetMulti$ be the category with objects ordinary sets, morphisms $f: X \to Y$ given by functions (i.e maps in the ordinary category of sets) $f: X \to \cP(Y)\setminus \emptyset$, and composition given by
    \[f\circ g(x) = \bigcup_{y \in g(x)} f(y)\]
    for $g: X \to Y, f: Y \to Z$.
    We think of the morphisms in $\SetMulti$ as \emph{multivalued functions}.
    See also \cite[Example~2.6]{markov_cats} for the analogous category of finite sets.
    The Cartesian product of sets gives it a symmetric monoidal structure, if we set \[(f \tensor g)(x_1,x_2) = f(x_1) \times g(x_2) \subset Y_1 \times Y_2,\] for $f:X_1 \to Y_1, g: X_2 \to Y_2$.
    This is a semicartesian monoidal structure, and we can give it a Markov category structure by setting
    $\cop{X}(x) = \{(x,x)\} \subset X \times X$.
    Then the deterministic morphisms are precisely those sending each point into a singleton set, and there is an obvious isomorphism of categories
    $\SetMulti_\detc \cong \sf{Set}$.
    Of course, $\sf{Set}$ actually has all infinite products, but they do not form Kolmogorov products, or even infinite tensor products in $\sf{SetMulti}$.
    To see this, consider the family $X_n = \{0,1\}$ for $n \in \bN$.
    A map $* \to \prod_{n\in \bN} \{0,1\} = X_\bN$ is just a subset of this infinite product.
    Given such a subset $U$, the finite marginals $* \to X_F$ for $F \subset \bN$ are the images of $U$ under the projection map.
    But these images do not determine $U$. For instance, consider the subset $U$ of those sequences containing at least one $1$.
    The image in $X_F$ is all of $X_F$ (since any finite sequence of $0$s and $1$s can be extended to one containing a $1$).
    But $U$ is not all of $X_\bN$.

@@

\begin{lemma}
	\label{aseqlemma}
	Let $\cC$ be a causal Markov category. Let $p : A \to X$ and $f,g: X \to Y$ be morphisms in $\cC$ with $f$ deterministic and such that
	\[
		\tikzfig{aseqlemma}
	\]
	Then $f =_{p\as} g$ (\Cref{defnasequal}).
\end{lemma}
\begin{proof}
    Consider the morphism
    \[
	    \tikzfig{aseqlemma1}
    \]
    \todo[inline]{E: I think here we need the second string going out of $g$ too, to fit with the form of \cref{causal_defn}? Although it's not critically important, the argument still works and we just marginalize that away too}.
    Applying the assumed equation together with determinism of $f$ shows that marginalizing over the first output gives the same morphism $A \to Y \otimes Y$ as marginalizing over the second output. Therefore, applying the causality property of \Cref{causal_defn} with $h_1$ and $h_2$ the two marginalizations, we can conclude the equality
    \[
	\tikzfig{aseqlemma2}
    \]
    Marginalizing this further over the middle output and truncating the redundant pieces proves the desired equation.
\end{proof}

@@

\todo[inline]{E: Not sure these examples (about top. spaces) are that great.. it seems like we do an inordinate amount of work to get the infinite tensor products, compared to the results. What do you think?}

@@

\begin{definition}
    Consider the hyperspace monad on $\Top$ defined in \cite[Definition~2.4]{hyperspace}.
    The Kleisli category associated to this monad is denoted $\Kl(H)$.
    Using \cite[Corollary~2.53]{hyperspace} and \cite[Proposition~3.1]{markov_cats}, this is a Markov category.
    Concretely, this category has the following structure:
    \begin{enumerate}
        \item The objects are topological spaces.
        \item A morphism $f: X \to Y$ consists of a continuous function $X \to HY$, where $HY$ denotes the set of closed subsets, equipped with the topology generated by subbasic opens of the form $\{V \subset Y \text{ closed} \mid V \cap U \neq \emptyset\}$ for $U \subset Y$ open.
        \item Composition is defined as in $\SetMulti$, by the closure of the union.
        \item The monoidal product is the Cartesian product of topological spaces, and $(f \tensor g)(x_1,x_2) = f(x_1) \times g(x_2)$.
        \item The comultiplication is defined by $\cop{X}(x) = \cl(\{(x,x)\})$
    \end{enumerate}
\end{definition}
\begin{proposition}
    $\Kl(H)$ has Kolmogorov products of any cardinality, given simply by the ordinary infinite product of topological spaces in the product topology.
\end{proposition}
\begin{proof}
    First we observe that $H(\prod_{j\in J} X_j)$ is the cofiltered limit of the spaces
    $H(\prod_{j\in F} X_j)$.
    To see this, note that the marginalizations induce a continuous map
    \[\alpha: H(\prod_{j\in J} X_j) \to \lim_{F \subset J} H(\prod_{j \in J} X_j)\]
    This is an injection - given distinct closed subsets $V, W \subset \prod X_j$, and let $x \in V \setminus W$.
    Then let $U \ni x$ be an open neighborhood of $x$ not intersecting $W$.
    Since $U$ is open in the product topology, we can find some finite subset $F \subset J$ so that 
    \[y = (y_j) \in U \Leftrightarrow y_j \in \pi_j(U) \text{ for $j\in F$}\]
    Then $\pi_F(x) \in \pi_F(U)$, which is disjoint from $\pi_F(W)$. Hence $\alpha(V) \neq \alpha(W)$.
